Higher-Order unification for free!

### 2024-04-16 paper/main.tex
wip

\noindent
We now illustrate the encoding of the three instances above as higher-order
logic programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises, and
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

@@

\noindent
Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work
since it uses the predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}. If we try to back-chain the rule
\ref{clause:r3} on the encoding of the goal \ref{goal:g} given below

@@

The meta language of choice is expected to provide 
an implementation of \Ue that satisfies
the following properties:
%~\ref{prop:correct-ml} and~\ref{prop:complete-ml}.
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
\end{gather}\todo{give a name and reuse it for \Uo in next sections}

@@

\begin{proposition}[\wellb-enforcing]\label{prop:w-enforcing}
  Given a term $s$ in \Fo{}, 
  $$
    \C{s}{t}{m}{l} \Rightarrow \wellb(t)
  $$
\end{proposition}

\begin{proposition}[compiler-useful]\label{prop:w-noncontra}
  Given two terms $s_1$ and $s_2$, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
  then 
  $$
    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
    t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-w}
  $$
\end{proposition}

\noindent
In other words the compiler outputs terms in \wellb, even if its
input is not. And if two terms can unify in the source language
then these terms also unify in the target language. 
Note that the property holds for any substitution, $\rho$ could be given by an
oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply
does not contradict \Eo. Also note that a compiler that
translate all terms to unification variables would satisfy this
property, but wound fail to verify \cref{prop:fidelity}.

@@

\begin{proposition}[Compilation round trip]\label{prop:comprt}
  % \elpiIn{compile F T [] M [] L [] S} then 
  % if \elpiIn{T} $\in$ S
  If $\C{s}{t}{m}{l}$, such that $l \in \linkStore$ and $m \in \mapStore$, if
  $\sigma = \{ A \mapsto t\}$ and $\mapping{X}{A}{n} \in \mapStore$
  then $\D{\sigma}{\mapStore}{\linkStore}{\rho}$ and
  $\rho X \Eo \rho s$.
\end{proposition}
\todo{explain no proof, llam link breaks}

% \noteD{Should also decompile links\dots}
% \begin{proof}[Proof sketch]
% Trivial since \mapStore is a bijection and 
% the terms are beta normal.
% some discussion about commit maybellam to be done later.
% \end{proof}

@@

\begin{lemma}[Properties of \Uo in \wellb]\label{prop:uniffow} The following properties hold for \Uo
  as implemented above: \todo{reuse statement from section 2.4}
\begin{gather}
  \unifcorrectX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}\\
  \unifcompleteX{\Eo}{\Uo}{t}{\wellb(\{t_1, t_2\})}
\end{gather}
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ X \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{definition}[$\wellb_\beta$]
$
\wellb_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
$
\end{definition}

\begin{lemma}[Properties of \Uo in $\wellb_\beta$]\label{prop:uniffowb} The following properties hold for \Uo
  where \todo{reuse 2.4}
  \begin{gather}
    \unifcorrectX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}\\
    \unifcompleteX{\Ee}{\Ue}{t}{\wellb_\beta(\{t_1, t_2\})}
  \end{gather}
\end{lemma}

@@

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybebeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_\beta(s)$  then $\:\wellb(t)$.
\end{lemma}
% \begin{proof}[Proof sketch]
%   problematic terms are mapped to uva by comp,
%   the problematic fapp node is gone.
%   if F x = t then  F = lam x.t[]
%   \todo{finish/maybe W enforcement is not needed here}
% \end{proof}
\noindent
In other word the compiler eliminates all subterms that are in \maybebeta.

\begin{theorem}[Fidelity in $\wellb_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb_\beta(\foUnifPb)$
\end{theorem}
\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
  \cref{prop:uniffowb}.
\end{proof}

@@

\begin{definition}[$\wellb_{\beta \eta}$]
  $
  \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
  $
\end{definition}

\begin{lemma}[$\wellb$-partial-enforcement]\label{lem:w-enforcement-maybeeta}
  $\forall s, \C{s}{t}{m}{l}$, if $\:\wellb_{\beta \eta}(s)$  then $\:\wellb(t)$.
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{lem:maybeeta}
\end{proof}

@@

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
  $\wellb_{\beta \eta}(\foUnifPb)$ and if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}
\todo{should we introduce W/eta as prev section W/beta?}

@@

% \begin{definition}[$\wellb_{\beta \eta \llambda}$]
%   $
%   \wellb_{\beta \eta}(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in \notllambda{}
%   $
% \end{definition}

\begin{lemma}[$\wellb$-enforcement]\label{lem:w-enforcement-maybellam}
  $\forall s, \C{s}{t}{m}{l}$, then $\:\wellb(t)$.
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{lem:w-enforcement-maybebeta,lem:w-enforcement-maybeeta} and the 
  rule above.
\end{proof}

@@

\begin{definition}[\progBetaLL]
  Let $\sigma$ be a substitution such that $\sigma t_1$ is a name $s$
  not occurring$s_1\dots s_n$. If $m = 1$, then $l$ is removed and \lhs is
  unified with $X_{s_1\ldots s_n~s}$.
  If $m > 1$, then $l$ is replaced by the link
  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n~s}\appsep t_2\dots t_m}$
  (where $Y$ is a fresh variable of arity $n+1$) and link
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n~x}}
  is added to \linkStore.

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH{}, then it disappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates. Finally \progBetaFail also guarantees termination since 
  it makes \elpiIn{progress} immediately fails.
%
  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  % \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

@@

\begin{theorem}[Fidelity in \Fo]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
  if \Uo{} fails outside \llambda.
\end{theorem}

>>>>

\noindent
We now illustrate a naive encoding of the three instances above as higher-order
logic programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises, and
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

@@

\noindent
Unfortunately this encoding of rule \ref{clause:r3} does not work
since it uses the predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}. If we try to back-chain the rule
\ref{clause:r3} on the encoding of the goal \ref{goal:g} given below

@@

The meta language of choice is expected to provide 
an implementation of \Ue that satisfies
the following properties:
%~\ref{prop:correct-ml} and~\ref{prop:complete-ml}.
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
\end{gather}

@@

\begin{proposition}[\wellb-enforcing]\label{prop:w-enforcing}
  Given two terms $s_1$ and $s_2$, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
  then 
  $$
    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
    t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-w}
  $$
\end{proposition}

\noindent
In other words the compiler outputs terms in \wellb, even if its
input is not.
Note that the property holds for any substitution, $\rho$ could be given by an
oracle and/or not necessarily be a most general one: in \wellb \Ue{} simply
does not contradict \Eo.

@@

\begin{lemma}[Compilation round trip]\label{prop:comprt} If
  \elpiIn{compile S T [] M [] _ [] _} then \elpiIn{decompile M T S}
\end{lemma}\noteD{Should also decompile links\dots}
% \begin{proof}[Proof sketch]
% Trivial since \mapStore is a bijection and 
% the terms are beta normal.
% some discussion about commit maybellam to be done later.
% \end{proof}

@@

\begin{lemma}[Properties of \Uo]\label{prop:uniffow} The following properties hold for \Uo:
\begin{gather}
  \unifcorrectX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}\\
  \unifcompleteX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}
\end{gather}
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

\begin{definition}[$\wellb/_\beta$]
$
\wellb/_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
$
\end{definition}

\begin{lemma}[Properties of \Uo]\label{prop:uniffowb} The following properties hold for \Uo
  where
  \begin{gather}
    \unifcorrectX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}\\
    \unifcompleteX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}
  \end{gather}
\end{lemma}

@@

\begin{lemma}[$\wellb/_\beta$-enforcement]\label{lem:w-enforcement-maybebeta}
  Given two terms $s_1$ and $s_2$ in \maybebeta, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
  then 
  $$
    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
    t_1 \Ue t_2 \mapsto \sigma
  $$ 
\end{lemma}
\begin{proof}[Proof sketch]
  problematic terms are mapped to uva by comp,
  the problematic fapp node is gone.
  if F x = t then  F = lam x.t[]
  \todo{finish/maybe W enforcement is not needed here}
\end{proof}

\begin{theorem}[Fidelity in $\wellb/_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb/_\beta(\foUnifPb)$
\end{theorem}
\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
  \cref{prop:uniffowb}.
\end{proof}

@@

@@

\begin{lemma}
  The introduction of \linketa guarantees \fullref{prop:w-enforcing}
  %\label{lemma:unif-wellb}
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{def:testmaybeeta}.\todo{shall we move the proof here?}
\end{proof}

@@

\begin{theorem}[Fidelity in \maybeeta]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow t \notin \notllambda$ and if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}
\todo{should we introduce W/eta as prev section W/beta?}

@@

@@

\begin{definition}[\progBetaLL]
  Let $\sigma$ be a substitution such that $\sigma t_1$ is a name $s$
  not occurring$s_1\dots s_n$. If $m = 1$, then $l$ is removed and \lhs is
  unified with $X_{s_1\ldots s_n~s}$.
  \todo{I changed this part, it is OK?}
  If $m > 1$, then $l$ is replaced by the link
  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n~s}\appsep t_2\dots t_m}$
  (where $Y$ is a fresh variable of arity $n+1$) and link
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n~x}}
  is added to \linkStore.

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH{}, then it disappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates. Finally \progBetaFail also guarantees termination since 
  it makes \elpiIn{progress} immediately fails.

  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  % \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

@@

\begin{theorem}[Fidelity with \linkbeta]
  The introduction of \linkbeta guarantees \fullref{prop:fidelity-recovery}
  if \Uo{} fails outside \llambda.
\end{theorem}
\todo{should we introduce W/llam as prev section W/beta and W/eta?}
